Large Language Models in Reinforcement Learning - A Comparison of Methods

### 2023-10-27 doc/paper.tex
Ad dmore to paper
Signed-off-by: Jonatan Hanssen <jonatan.hanssen@outlook.com>

Reinforcement Learning (RL) algorithms suffer when rewards are sparse and the state-action space is large. Even tasks which seem relatively simple can prove intractable if the completion of the task requires subtasks to be completed first, or if the reward only comes when the entire task has been completed. In such cases, random exploration is unlikely to lead the agent to discover a solution to the problem. The apparent simplicity of such problems is often due to human intuition, which allow us to quickly see possible solutions to a diverse set of problems. Large Language Models (LLMs) are trained on large corpora of human written text, and have been shown to capture parts of this intuition in many tasks \cite{sparks}. In recent years, LLMs have been succesfully used to aid RL agents in more efficient exploration, and have allowed them to solve problems which previous methods have been unable to \cite{omni} \cite{ellm}. We compare different methods for integrating an LLM into the Deep Learning RL algorithm Proximal Policy Optimization (PPO), and compare their efficiency against each other. We find that BEST METHOD gives the best sample efficiency, outperforming normal PPO by PERCENTAGE in METRIC.

@@

One of the central challenges of reinforcement learning is that rewards are often both extremely rare and delayed, which means that optimizing a reinforcement learning algorithm requires significant trial and error \cite[423]{brunton}. Furthermore, in many problems the state-action spaces are enormous, meaning that uniform exploration is unlikely to find good solutions in a reasonable amount of time. Many methods have been developed to deal with this issue, for example by introducing auxilliary reward functions that use domain knowledge to reward actions which are considered good, or which reward the agent for learning novel skills. However, hand picking which actions to reward can lead to imitation rather than optimal behaviour, and novelty is not always useful \cite[1]{ellm}. In recent years, LLMs have shown remarkable capabilities in problem solving and planning \cite{sparks}, qualities which traditional RL agents often lack. Thus, a new area of research has emerged, which attemps to use the vast amount of human knowledge encoded in these models to improve the performance of reinforcement learning algorithms \cite{survey}. These LLM assisted RL agents have been able to outperform state-of-the-art RL methods in many problems \cite{omni} \cite{ellm} \cite{idm}.

@@

RL informed by natural language is a relatively new field, where research before 2018 have been limited to relatively small corpora or synthetic language \cite[1]{survey}. With the introduction of Large Language Models, many new possibilities have opened up, and in recent years there have been many successful integrations of these models into traditional RL \cite{omni} \cite{ellm} \cite{idm}.

@@

However, rather than doing so through directly influencing action probabilities in the policy, the paper suggests rewarding the agent if it takes actions that are semantically similar to actions suggested by an LLM. At certain timesteps, the LLM is prompted with a state observation from the agent, and is asked to suggest actions for the agent to take. If the agent then takes actions which are semantically similar to the suggested actions, it recieves a reward. Thus, the proposed algorithm makes use of LLMs in two ways: first to generate what the LLM perceives as desirable actions, then later to assess similarity between a natural-language description of the agents action and the previously suggested actions. The first use case in particular utilizes the presence of human intuition in LLMs, as a human would often be able to intuitively figure out what is desirable to achieve in some environment.

@@

We select NUMBER different Minigrid environments; DoorKey, Unlock, CRAZYENV. Example instances of these environments can be found in the appendix. For each environment we test our three different models; PPO with no LLM integration, PPO with LLM reward shaping and PPO with LLM policy integration. All methods use the same hyperparameters for the PPO, which can be found the appendix. For each of these methods, we run PPO for 2000 episodes of 1024 steps, and track the reward\footnote{The true environment reward, not including pseudo-rewards given by the LLM} received for each one. We do this eight times per method, and compare the average results against eachother.

@@

In accordance with the NeurIPS Code of Ethics ***ADD REFERENCE HERE***, we outline ethical considerations related to this project, covering potential consequences from the research, and how potential risks may be mitigated. Considering that our research does not involve participants nor sensitive data, concerns related to how such factors are treated will not be discussed.

@@

>>>>

Reinforcement Learning (RL) algorithms suffer when rewards are sparse and the state-action space is large. Even tasks which seem relatively simple can prove intractable if the completion of the task requires subtasks to be completed first, or if the reward only comes when the entire task has been completed. In such cases, random exploration is unlikely to lead the agent to discover a solution to the problem. The apparent simplicity of such problems is often due to human intuition, which allow us to quickly see possible solutions to a diverse set of problems. Large Language Models (LLMs) are trained on large corpora of human written text, and have been shown to capture parts of this intuition in many tasks. In recent years, LLMs have been succesfully used to aid RL agents in more efficient exploration, and have allowed them to solve problems which previous methods have been unable to. We compare different methods for integrating an LLM into the Deep Learning RL algorithm Proximal Policy Optimization (PPO), and compare their efficiency against each other. We find that BEST METHOD gives the best sample efficiency, outperforming normal PPO by PERCENTAGE in METRIC.

@@

One of the central challenges of reinforcement learning is that rewards are often both extremely rare and delayed, which means that optimizing a reinforcement learning algorithm requires significant trial and error \cite[423]{brunton}. Furthermore, in many problems the state-action spaces are enormous, meaning that uniform exploration is unlikely to find good solutions in a reasonable amount of time. Many methods have been developed to deal with this issue, for example by introducing auxilliary reward functions that use domain knowledge to reward actions which are considered good, or which reward the agent for learning novel skills. However, hand picking which actions to reward can lead to imitation rather than optimal behaviour, and novelty is not always useful \cite[1]{ellm}. In recent years, LLMs have shown remarkable capabilities in problem solving and planning \cite{sparks}, qualities which traditional RL agents often lack. Thus, a new area of research has emerged, which attemps to use the vast amount of human knowledge encoded in these models to improve the performance of reinforcement learning algorithms \cite{survey}. These LLM assisted RL agents have been able to outperform state-of-the-art RL methods in many problems \cite{omni, ellm, idm}.

@@

RL informed by natural language is a relatively new field, where research before 2018 have been limited to relatively small corpora or synthetic language \cite[1]{survey}. With the introduction of Large Language Models, many new possibilities have opened up, and in recent years there have been many successful integrations of these models into traditional RL \cite{omni, ellm, idm}.

@@

However, rather than doing so through directly influencing action probabilities in the policy, the paper suggests rewarding the agent if it takes actions that are semantically similar to actions suggested by an LLM. For example, the LLM might recommend the agent to "cut down a tree". If the agent performs the "chop" action at a tree, this action can be described as "chop tree". Using a language model, we can create encodings of these sentences and compare their similarity mathematically. At certain timesteps, the LLM is prompted with a state observation from the agent, and is asked to suggest actions for the agent to take. If the agent then takes actions which are semantically similar to the suggested actions, it receives a reward. Thus, the proposed algorithm makes use of LLMs in two ways: first to generate what the LLM perceives as desirable actions, then later to assess similarity between a natural-language description of the agents action and the previously suggested actions. The first use case in particular utilizes the presence of human intuition in LLMs, as a human would often be able to intuitively figure out what is desirable to achieve in some environment.

@@

We select NUMBER different Minigrid environments; DoorKey, OTHERENV, OTHERENV. Example instances of these environments can be found in the appendix. For each environment we test our three different models; PPO with no LLM integration, PPO with LLM reward shaping and PPO with LLM policy integration. All methods use the same hyperparameters for the PPO, which can be found the appendix. For each of these methods, we run PPO for 2000 episodes of 1024 steps, and track the reward\footnote{The true environment reward, not including pseudo-rewards given by the LLM} received for each one. We do this eight times per method, and compare the average results against eachother.

@@

In accordance with the NeurIPS Code of Ethics \cite{ethics}, we outline ethical considerations related to this project, covering potential consequences from the research, and how potential risks may be mitigated. Considering that our research does not involve participants nor sensitive data, concerns related to how such factors are treated will not be discussed.

@@

% \begin{table*}[h]
% \caption{PPO hyperparameters}
% \begin{center}
% \label{hyperparams}
% \begin{tabular}{c | l l l l l l l l} 
% % & & \\ % put some space after the caption
% % \hline
% Parameter & Gamma & Steps/episode & Epochs/episode & Batches/epoch & Clip & Value loss coeff. & Entropy coeff. & Learning rate \\
% \hline
% Value & 0.99 & 1024 & 10 & 8 & 0.2 & 0.3 & 0 & 3.0e-4 \\
% \hline
% \end{tabular}
% \end{center}
% \end{table*}
\begin{table*}[h]
\caption{PPO hyperparameters}
\begin{center}
\label{hyperparams_transposed}
\begin{tabular}{c | c}
Parameter & Value \\
\hline
Gamma & 0.99 \\
Steps/episode & 1024 \\
Epochs/episode & 10 \\
Batches/epoch & 8 \\
Clip & 0.2 \\
Value loss coeff. & 0.3 \\
Entropy coeff. & 0 \\
Learning rate & 3.0e-4 \\
\end{tabular}
\end{center}
\end{table*}
